---
title: "week1_notes"
author: "ABZ"
date: "May 22, 2020"
output: html_document
---

``` {r}
myfunction <- function() {
  x <- rnorm(100)
  print(x)
  mean(x)
}

second <- function(x) {
  x + rnorm(length(x))
}
```

## Missing Values
NaN is NA but NA is not NaN. NA's can have classes

``` {r}
x <- c(1, 2, NA, 20, 3)
is.na(x)
is.nan(x)

x <- c(1, 2, NaN, NA, 4)
is.na(x)
is.nan(x)
```

## Data frame
stores tabular data. Every element of the list has to be of the same length. Unlike a matrix, each 
column (element) can be a different class. Special attributes, row.names - each row has a name. 
Created by calling read.table() or read.csv(). data.frame() builds a dataframe from scratch.

``` {r}
x <- data.frame(foo = 1:4, bar = c(T, T, F, F))
x
nrow(x)
ncol(x)
```

## Names
R objects can have names! No name by default but can assign.

``` {r}
x <- 1:3
names(x)
names(x) <- c("foo", "bar", "norf")
x
names(x)
```

Lists can also have names!

``` {r}
x <- list(a = 1, b = 2, c = 3)
x
```

matrices can have names - dimnames()

``` {r}
m <- matrix(1:4, nrow = 2, ncol = 2)
dimnames(m) <- list(c("a", "b"), c("c", "d"))
m
```

## Reading Tabular Data

text files that returns a dataframe

read.table()
read.csv()

Note: read.table reads in characters as factor as a default. This could be problematic for text
analytics, so setting stringasfactor = FALSE may be desirable. Read.table returns a data frame
Note read.table() is the same as read.csv() except that read.csv is comma deliminated by default


## Reading in Larger Tables
READ THE HELP PAGE FOR read.table()!!
Make sure enough RAM is allocated to store dataset
Use the colClasses argument. set class to "numeric" makes R assume that all columns are numeric.
This makes R run MUCH faster, often by x2!
A quick and ditry way to figure out the classes of wach column is the following:

``` {r}
initial <- read.table("datatable.txt", nrows = 100)
classes <- sapply(initial, class)
tabAll <- read.table("datatable.txt", colClasses = classes)
```

set nrows will not help R run faster but will help with memorty usage.

## Textual Data formats
dput - can only be used on a single R object

``` {r}
y <- data.frame(a = 1, b = "a")
dput(y)
dput(y, file = "y.R")
new.y <- dget("y.R")
new.y
```

#dump 
similar to dput but can be applied to several objects

``` {r}
x <- "foo"
y <- data.frame(a = 1, b = "a")
dump(c("x", "y"), file = "data.R")
rm(x, y)
source("data.R")
y
x
```

#Reading lines of a text file - readLines()

``` {r}
con <- url("http://jhsph.edu", "r")
x <- readLines(con)
head(x)
```

##Subsetting objects in R

[] - always returns an object of the same class as the original, can be used to select more than 
one element (there is one excption)

[[]] - is used to extract elements of a list or a data frame, it can only be used to extract a 
single element and the class of the returned object will not necessarily be a list or data frame

$ - is used to extract elements of a list or data frame by name, semantics are similar to that 
of [[]]

# Examples, numeric indexes

```{r}
x <- c("a", "b", "c", "c", "d", "a")
x[1] # exctracts the first element of x 
x[2] # extracts the second element of x
x[1:4] # extracts the first four elements of x
```

# Examples of logical index

``` {r}
x[x > "a"]
u <- x > "a"
x[u]
```

##Subsetting lists

``` {r}
x <- list(foo = 1:4, bar = 0.6)
x[1]       #list that contains 1 through 4
x[[1]]     #sequence of 1 through 4
x$foo

x$bar      #give element that is associate with the name "bar"
x[["bar"]] 
x["bar"]   #Note using the name is nice because you do not have to remember the numeric index
```

extract multiple elements form a list

``` {r}
x <- list(foo = 1:4, ba = 0.6, baz = "hello")
x[c(1 ,3)] # returns list of foo and baz. Can't use [[]]
```

but we can use[[]] for other things:

``` {r}
name <- "foo"
x[[name]]
x$name # element 'name' doesn't exist!
x$foo # element 'foo' does exist
```

When subsetting nested elements of a list. The [[]] can take an integer sequence

``` {r}
x <- list(a = list(10, 12, 14), b = c(3.14, 2.81))
x[[c(1, 3)]]
x[[1]][[3]]
x[[c(2, 1)]]
```

## Subestting Matrices

``` {r}
x <- matrix(1:6, 2, 3)
x[1,2]
x[2,1]
```

Indices can also be missing

``` {r}
x[1, ] #first row of matrix
x[, 2] #second column of matrix
```

NOTE: by default, when a single element of a matrix is retrieved, it is returned as a vector of 
length 1 rather than a 1x1 matrix. This behavior can be altered by setting drop = FALSE

``` {r}
x[1, 2]
x[1, 2, drop = FALSE]
x[1, ]
x[1, , drop = FALSE]
```

##Removing NA values

``` {r}
x <- c(1, 2, NA, 4, NA, 5)
bad <- is.na(x)
x[!bad]
```

more complicated case from multiple elements

``` {r}
x <- c(1, 2, NA, 4, NA, 5, 6)
y <- c("a", "b", NA, "d", NA, "f", NA)
good <- complete.cases(x, y)
good
x[good]
y[good]
```

Here is an example

``` {r}
airquality[1:6, ]
good <- complete.cases(airquality)
airquality[good, ][1:6, ]
```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

